\SdM{Gaia identifies VFTS682 as the most massive, dynamically ejected runaway star known to date}

### 2018-05-11 VFTS682.tex
small

\documentclass{aa}
\usepackage{graphicx}
\usepackage[varg]{txfonts}
\usepackage[%draft, 
colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{amsmath}
\usepackage[usenames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
\usepackage{chngpage}
\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
\usepackage{forest}
\usepackage{tikz-qtree}
\usetikzlibrary{shadows,trees}

%Selma's comments
\definecolor{Wildstrawberry}{rgb}{1.0, 0.26, 0.64}
\newcommand{\SdM}[1]{{{\color{Wildstrawberry}\bf{#1}}}}
\newcommand{\newtext}[1]{{\color{ForestGreen}\bf{#1}}}

\renewcommand{\labelitemii}{$\bullet$}
\newcommand{\kms}{{\,\mathrm{km\ s^{-1}}}}
\newcommand{\Msun}{{\mathrm{M}_\odot}}

@@

\title{VFTS682: a confirmed dynamical ejection?}

\author{M.~Renzo\inst{1} \and S.~E.~de~Mink\inst{1} \and \SdM{D.J. Lennon et al.} \todo{etc...}%\and
  % S.~Justham\inst{2,3} \and A.~de~Koter\inst{1} \todo{etc..order to be decided}
  .}

\institute{{Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands}
  % \and{School of Astronomy \& Space Science, University of the Chinese
  %   Academy of Sciences, Beijing 100012, China}
  % \and{National Astronomical Observatories, Chinese Academy of
  %   Sciences, Beijing 100012, China}
}

\offprints{M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\abstract{}

@@

How do massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and they remain enshrouded in their parent cloud during the formation
process.  Major progress has been made on the theoretical side,  \citep[e.g.][]{Kuiper+2015Rosen+2016}, but the simulations are computationally very expensive. They can typically not yet resolve the length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[][]{sana:12,sana:17}.  Understanding massive star formation, and its possible dependence on environment and metallicity, is crucial for understanding the role massive stars play within their host galaxies as sources of feedback, but also for understanding the transients that mark their death and the compact remnants they leave behind.  This is therefore a key question for the present and upcoming transient surveys \citep[e.g., LSST, BlackGem, LIGO/Virgo O3][]{}\todo{ref} which  will reveal transients associated to massive stars
evolution and death.

@@

Our results indicate that VFTS 682 is a bona fide runaway star.  The direction an magnitude of the velocity vector is full consistent with dynamical ejection from R136.  Its inferred age is consistent with the inferred travel time. This means that the hypothesis that VFTS 682 is formed in relative isolation is rejected.  This makes  VFTS 682 the most massive runaway star known to date.

}

% make sure to cite this in the conclusion \cite{fujii:11, banerjee:12}
\todo{missing: R136 too young for CC, cite age $\leq2$\,Myr \citep[][]{crowther:10,sabbi:12}

\section{Gaia DR2 data selection}
\label{sec:sample}

VFTS682 is labeled in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}} with the
source id 4657685637907503744. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}. Its reported G-band
magnitude is 15.65, cf. the V-band magnitude of 16.08
\citep[][]{evans:11, bestenlehner:11}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are trustworthy. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/main_plot}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region, starting from the present day position of
    the star. The semi-transparent arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards to illustrate the uncertainty on the origin of the
    star. Blue stars
    are a subset of those we use to define the local rest frame. \todo{load color picture on background}}
  \label{fig:main}
\end{figure}

\section{Discussion}
\label{sec:discussion}

@@

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\item another important thing missing: with vsini(i)<200, completely average, there is no reason to think it's a CHE, contrary to the way they estimate the age in Bestenlehner:11
\end{itemize}

\bibliographystyle{aa}
\bibliography{bibliography/vfts682}

\begin{acknowledgements}
  We are grateful to S.~Torres for extremely useful discussions on the
  Gaia DR2 dataset, and to M.~C.~Ramirez-Tannus for invaluable help.
  The background image of \Figref{fig:main} is based on observations
  made with ESO Telescopes at the La Silla Observatory under programme
  ID 076.C-0888, processed and released by the ESO VOS/ADP group.
  This research made use of APLpy, an open-source plotting package for Python \citep[][]{robitaille:12}.
\end{acknowledgements}
\end{document}

>>>>

\documentclass[apjl,twocolumn]{emulateapj}

\usepackage{graphicx}
\usepackage[varg]{txfonts}

\usepackage[pagebackref,breaklinks,colorlinks,citecolor=blue]{hyperref}        
\renewcommand*{\backref}[1]{[#1]}  % gives neat back references in the pdf
%\usepackage[%draft, 
%colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{aas_macros}
%\usepackage{amsmath} % clashes with emulateapjj for some reaons
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
%\usepackage{chngpage}
%\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
%\usepackage{forest}
%\usepackage{tikz-qtree}
%\usetikzlibrary{shadows,trees}

%Selma's comments
\definecolor{Wildstrawberry}{rgb}{1.0, 0.26, 0.64}
\newcommand{\SdM}[1]{{{\color{Sepia}{#1}}}}
\renewcommand{\SdM}[1]{{{{#1}}}}

\newcommand{\newtext}[1]{{\color{ForestGreen}\bf{#1}}}

\renewcommand{\labelitemii}{$\bullet$}
\newcommand{\kms}{{\,\mathrm{km\ s^{-1}}}}
\newcommand{\Msun}{{\,\mathrm{M}_\odot}}
\newcommand{\Lsun}{{\,\mathrm{L}_\odot}}
\newcommand{\masyr}{{\,\mathrm{mas}\,\mathrm{yr}^{-1}}}

@@

\title{ \SdM{Gaia identifies VFTS682 as the most massive, dynamically ejected runaway star known to date}}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, \SdM{D.J. Lennon$^{2}$ et al.} \todo{... add coauthors...}%\and
  % S.~Justham\inst{2,3} \and A.~de~Koter\inst{1} \todo{etc..order to be decided}
  .}

\affil{$^{1}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{2}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78, E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
    $^{3}$ ... 
  % \and{School of Astronomy \& Space Science, University of the Chinese
  %   Academy of Sciences, Beijing 100012, China}
  % \and{National Astronomical Observatories, Chinese Academy of
  %   Sciences, Beijing 100012, China}
}

\thanks{M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
\SdM{ The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous work spectroscopic work identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 32 pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, lead to speculations whether such extreme stars can form in relative isolation. 
%
We use Gaia's unprecedented proper motion measurements to investigate the kinematics of this object. We show that VFTS 682 is a bonafide runaway star moving with a relative velocity of about $60\kms$ away from the star cluster R136.  The inferred age, distance, and the magnitude and direction of the velocity are consistent a dynamical ejection scenario from the cluster. 
%
If correct, this would make VFTS 682 the most massive dynamically ejected runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above the canonical upper mass limit inside the cluster, the internal dynamics in dense star clusters and the final fate of this object. }

\end{abstract}

@@

How do massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and they remain enshrouded in their parent cloud during the formation
process.  Major progress has been made on the theoretical side,  \citep[e.g.][]{Kuiper+2015,Rosen+2016}, but the simulations are computationally very expensive. They can typically not yet resolve the length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[][]{sana:12,sana:17}.  Understanding massive star formation, and its possible dependence on environment and metallicity, is crucial for understanding the role massive stars play within their host galaxies as sources of feedback, but also for understanding the transients that mark their death and the compact remnants they leave behind.  This is therefore a key question for the present and upcoming transient surveys e.g., LSST and LIGO/Virgo,
%\citep[][]{}\todo{ref} . Blackgem is still hypothetical and a very tiny project compared to the others.  
which  will reveal transients associated to massive stars
evolution and death.

@@

Our results indicate that VFTS 682 is a bona fide runaway star.  The direction an magnitude of the velocity vector is fully consistent with dynamical ejection from R136.  Its inferred age is consistent with the inferred travel time. This means that the hypothesis that VFTS 682 is formed in relative isolation is rejected.  This makes VFTS 682, with an inferred present day mass well above a hundred solar masses, the most massive runaway star known to date. 
}

\todo{(1. Make sure we cite and disucss fujii:11, banerjee:12 in discussion. 2. Make link to dynamical interaction proposed to form BHBH systems. }

\section{Observations}
\label{sec:sample}
\SdM{
In this section we summarize the remarkable properties derived for VFTS682 \Secref{data:vfts683} in earlier studies and we discuss how we Gaia data in 
\Secref{data:gaia}.

\subsection{ Overview of the stellar parameters of VFTS682 derived in previous studies \label{data:vfts683}}

The star VFTS682  \citep{Evans+2011}, located at  RA 05:38:55.510  Dec -69:04:26.72 \todo{format properly}
was originally classified as a young stellar object \citep{Gruendl+2009} based on its mid-infrared excess.  \citet{Evans+2011} classified the object a Wolf-Rayet star using spectroscopic observations taken a part of the VLT-FLAMES Tarantula survey.  The spectra available covered $\lambda$4000--7000 and were taken at multiple epochs. \citep{Bestenlehner+2011} studied this object in more detail used and looked for radial velocity variations. This allowed them to exclude the presence of a close companion with high confidence.

\citep{Bestenlehner+2011} further analyzed the spectra using the non-LTE model atmosphere code CMFGEN \citep{Hillier+1998} to derive the stellar parameters and surface abundances. They derive a find a peculiar extinction ($R_V \sim 4.7$), leading to a surprisingly high luminosity $\log_{10} L (\rm{L_\odot}) =  6.5\pm0.2$. Bayesian analysis with the Bonsai code \citep{Schneider+2017} against  evolutionary models by \citet{Brott+2011} and \citet{Kohler+2015} leads to an estimated present-day mass of  $138^{+28}_ {-16}\Msun$, an age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial masse of $150^{+29}_{-17}\Msun$.  This places around the ``canonical upper limit'' of $150\Msun$  by \citet{Figer2005}.

The star is very similar to R136a3, located inside the star cluster R136, \citep{Crowther+2010}  for which \citet{Crowther+2016} report a current mass estimate of $180^{30}_{-30\Msun}$. The R136 cluster hosts two further very massive WN5h stars R1361 and R136a2, whose estimated current masses are even higher.   However, VFTS682 stands out by its location a projected offset of 32\,pc from the center of the cluster.

Further worth noticing is the variability of the star.  \citep{Bestenlehner+2011} discuss the Optical Gravitational Lensing Experiment (OGLE-III) light curves \citep{Udalski+2008} and show a variability in the V-band of the 10\% level on a timescale of years.  The comment that this is unusual for Wolf-Rayet and more reminiscent of Luminous Blue Variables.

\todo{Report mass loss rate.  Report LOS velocity.}

%\citep{Parker1993}

\subsection{Kinematic data from Gaia DR2  \label{data:gaia}}

The Gaia Data Release 2, which became available on 25 April 2018, provides the five-parameter astrometric solution (positions, parallaxes, and proper motions) for more than a bilion sources \citep{Gaia-Collaboration+2018}.  The median uncertainty in the proper motion components the corresponding uncertainties are $0.05\,\masyr$ for bright sources  ($G<14$ mag) and $0.2\masyr$ at $G=17\,\mathrm{mag}$ \citep{Lindegren+2018}.  The uncertainties quoted above to $xx\kms$ and $yy\kms$ respectively.  This unprecedented accuracy allows the identification of bright runaway stars at the distance to the LMC, 50\,kpc.

VFTS682 is identified with the gaia source} with id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia reports a G-band magnitude is 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are trustworthy. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/main_plot}  
  \caption{The red cross indicates the position of VFTS682. Blue stars
    are those we use to define the generic ``surroundings'', while the
    green dots indicate the stars we use to probe the core of R136,
    magnified in the inset. The red arrow shows the direction of the
    proper motion of VFTS682 relative to R136, and the gray shade
    indicates the uncertainty in the direction. \todo{load color picture on background}}
  \label{fig:main}
\end{figure}

\section{Discussion}
\label{sec:discussion}
\todo{Selma: Should add caveat that it may be a spurious detection.  That reconfirmation in following epochs is warranted. }

@@

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}

\todo{Merge Selma's my\_bib bibtex file with Mathieu's and prevent double refs. }
\bibliographystyle{aa}
\bibliography{bibliography/vfts682,my_bib}

\begin{acknowledgements}
  We are grateful to S.~Torres for 
  %extremely 
  very useful discussions on the
  Gaia DR2 dataset, and to M.~C.~Ramirez-Tannus for invaluable help.
  \SdM{SdM has received funding under the European Unions Horizon 2020 research and innovation programme from the European Research
Council (ERC) (Grant agreement No. 715063).
   This work has made use of data from the ESA space mission Gaia (http://www.cosmos.esa.int/gaia), processed by the Gaia Data Processing and Analysis Consortium (DPAC, http://www.cosmos.esa.int/web/gaia/dpac/consortium). Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the Gaia Multilateral Agreement. }
  The background image of \Figref{fig:main} is based on observations
  made with ESO Telescopes at the La Silla Observatory under programme
  ID 076.C-0888, processed and released by the ESO VOS/ADP group.
  This research made use of APLpy, an open-source plotting package for Python \citep[][]{robitaille:12}.
\end{acknowledgements}
\end{document}
